Best Practices for Multiple Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-05-15 paper/manuscript.tex
Adding an illustration of the timing diferences.
add first draft of adjusted networks section; still need to add citations

General simulation planning can be conceptualized as a three-component framework. Starting, the process involves identifying all mappings of the involved molecules (e.g. atom mappings), which define the structural relationships and transformations between molecule A and molecule B. Subsequently, these changes, along with their shared structural elements, are evaluated by estimating the difficulty of the transformations usually with predefined scoring functions. The final step entails constructing the perturbation graph using the scores from the prior step, which serves as a fundamental tool setting up a simulation campaign.

\paragraph{Atom mappings}
For minor molecular transformations, like in most large-scale approaches, single-coordinate set approaches are typically utilized due to their relatively low computational cost. These methods necessitate atom mappings to ensure their effective application. In large-scale implementations, the efficiency and robustness of automated mapping techniques are critical factors in these endavors.

%SIS
Subgraph isomorphism solver (SIS)-based maximum common substructure (MCSS) search methods currently represent the most widely used approach for addressing atom mapping challenges, at the current timepoint. Several cheminformatics tools implementing these solutions are readily available, including pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, and SMArt~\cite{petrov2021perturbation}.

These methods exhibit high accuracy in processing topological 2D information. However, hybrid approaches such as pmx~\cite{gapsys2015pmx} and LOMAP~\cite{liu2013lead}, which incorporate 3D structural data, enhance robustness in stereochemical considerations and molecular alignment, requiring well-prepared input coordinates. Regarding computational efficiency, SIS solvers must contend with the inherent complexity of solving an NP-complete problem~\cite{raymond2002maximum, Duesbury2018Comparison}. The use of an initial MCSS seed can mitigate this computational burden~\cite{raymond2002maximum}.

%Geometrz
An alternative to SIS-based MCSS methods is geometric bipartite matching for atom mapping, as implemented in tools such as pmx~\cite{gapsys2015pmx} and Kartograf~\cite{ries2024kartograf}. These algorithms enable rapid mappings due to their reduced computational complexity ($O(N) = N^3$), provided that well-aligned input geometries are available~\cite{ries2024kartograf}, through e.g. restraint-docking or prior-alignment. To accommodate topological variations, Kartograf applies a subsequent filtering step following the 3D mapping process.

% Short discussion of robustness and speed
Both mapping approaches present distinct advantages and limitations, each associated with specific prerequisites. Optimizing the robustness, and ignoring computational effort (e.g. mitigated by parallelization), it may be beneficial to employ both methods and implement a scoring mechanism to evaluate their performance. Most mappings are expected to be highly similar, this strategy could enhance robustness in the mapping process.

In terms of computational efficiency, the choice between methods largely depends on input conditions. If molecular structures are pre-aligned, the geometric approach offers a substantial speed advantage. Conversely, when ligands lack predefined 3D coordinates, the SIS-based method provides a easy solution, allowing coordinate generation at a later stage. This nuanced selection process ensures optimal performance based on the specific characteristics of the molecular system under investigation.

In combination with a suitable mapping scoring function the next steps is to build a pertubation map.

\paragraph{Perturbation maps}
\tocsectioncomment{Perturbation maps paragraph: @Mary Pitman/ @Jenke Scheen/ OpenFE team to update discussion of perturbation maps?}
Based on the input ligand series, a perturbation map or network can be planned. Recent heuristics have shown the more connected the perturbation network the better. However, there is a way to optimize network structure while minimizing the number of perturbations that need to be computed reducing the resulting computational cost~\cite{yang2020optimal,xu2019optimal}. Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or to deal with perturbations that would otherwise result in large numbers of atoms being inserted or deleted. Some commercial tools have good underlying heuristics but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.

@@

In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial free energy ranking, or in cases where perturbations would be rather large, a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used. However, adding redundancy into the network means that a better error analysis can be carried out by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks}\textbf{B}.

Methods from experimental design have been applied to the construction of the perturbation maps. Yang et al.~\cite{yang2020optimal} showed how to optimize the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance. Xu~\cite{xu2019optimal} showed how to optimize the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations. Both approaches lead to substantial reduction in the statistical error of the estimated free energies.

>>>>

\paragraph{Perturbation maps}
\tocsectioncomment{Perturbation maps paragraph: @Mary Pitman/ @Jenke Scheen/ OpenFE team to update discussion of perturbation maps?}
For RBFE campaigns with more than two ligands in the input series a \textit{perturbation map} or \textit{-network} must be generated: this is simply a set of data that describe which transformations will be executed between the ligands (referred to as \textit{edges}). \cite{MOBLEY LOMAP} Heuristics have shown the more connected the perturbation network the higher the free energy prediction accuracy versus experimental measures. Running a fully-connected network (i.e. simulate transformations between \textit{all} ligands in the campaign) is typically considered too computationally expensive and may not be suited optimally with current estimators such as MLE (\ref{SECTION MLE}).\cite{SCHEEN ET AL} There exist several methods for optimization of network structure while minimizing the number of perturbations that need to be computed, reducing the resulting computational cost at a minimal decrease in accuracy ~\cite{yang2020optimal,xu2019optimal}. Sometimes - depending on the RBFE implementation - the introduction of intermediate ligands that are not part of the original congeneric series can be helpful for avoiding challenging transformations such as ring breaking or transformations that would otherwise result in large numbers of atoms being inserted or deleted.\cite{LINDE PAPER} Some commercial tools have well-calibrated underlying heuristics to plan these network designs but may fail with complicated input datasets, thus requiring manual user validation, in particular when dealing with chiral compounds.

@@

In some cases, when dealing with for example very large datasets that would benefit from rougher initial free energy ranking or in cases where perturbations would be rather large, a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used. However, adding redundancy into the network means that a better error analysis can be carried out by analyzing cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks}\textbf{B}.

Methods from experimental design have been applied to the construction of perturbation maps. Yang et al.~\cite{yang2020optimal} showed how to optimize the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance. Xu~\cite{xu2019optimal} showed how to optimize the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations. Both approaches lead to substantial reduction in the statistical error of the estimated free energies.
